\texttt{MSnbase}, efficient and elegant R-based processing and visualisation of raw mass spectrometry data

### 2020-07-24 msnbase2.Rnw
further updates/cleaning of text

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. We focus on the new \textit{on-disk} infrastructure, that
  allows the handling of large raw mass spectrometry experiment on
  commodity hardware and illustrate how the package is used for
  elegant data processing, method development, and visualisation.
\end{abstract}

@@

Mass spectrometry is a powerful technology to assay chemical and
biological samples. It is used in routine applications with well
characterised \textcolor{red}{protocols} \textcolor{red}{such as
  diagnostics}, as well as a development platform,
\textcolor{red}{with the aim} to improve on existing \sout{methods}
\textcolor{red}{protocols} and devise new ones. The complexity and
diversity of mass spectrometry yield data that is itself complex and
often times of considerable size, that require non trivial processing
before producing interpretable results. \sout{This is particularly
  relevant, and can} \textcolor{red}{The complexity and size of these
  data} constitute a significant challenge for \sout{method}
\textcolor{red}{protocol} development: in addition to the development
of sample processing and mass spectrometry methods
\textcolor{red}{that yield the raw data}, \textcolor{red}{it is
  essential} \sout{there is a need} to process, analyse, interpret and
assess these new data to demonstrate the improvement in the technical,
analytical and computational workflows.

Practitioners have a diverse catalogue of software tools to explore,
process and interpret mass spectrometry data at their disposal. These
range from low level software libraries that are aimed at programmers
to enable development of new applications, to more user-oriented
applications with graphical user interfaces which provide a more
limited set of functionalities to address a defined scope. Examples of
software libraries include Java-based jmzML~\cite{Cote:2010} or
C/C++-based ProteoWizard~\cite{Chambers:2012}.
\sout{ProteomeDiscoverer (Thermo Scientific)}
\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
most widely used user-centric applications.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata (Figure \ref{fig:raw}). Each spectrum is
composed of m/z values and associated intensities. The metadata are
represented by a \textcolor{red}{single} table with variables along
the columns and each row associated to a spectrum. Among the metadata
available for each spectrum, there are MS level, acquisition number,
retention time, precursor m/z and intensity (for MS level 2 and
above), and many more. \texttt{MSnbase} \textcolor{red}{relies on the
  \texttt{mzR} package~\cite{Chambers:2012} to import raw mass
  spectrometry data from one of the many community-maintained open
  standards formats (mzML, mzXML, mzData or ANDI-MS/netCDF) and}
provides a rich interface to manipulate such objects. The code chunk
below illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{./figure/raw.png}
  \caption{Schematic representation of what is referred to by
    \textit{raw data}: a collection of mass spectra and a table
    containing spectrum-level annotations along the lines. Raw data is
    contained in one of the many community-maintained open standards
    formats (mzML, mzXML, mzData or ANDI-MS/netCDF). }
  \label{fig:raw}
\end{figure}

@@

In the following sections, we describe \sout{\texttt{MSnbase}'s
  ability to efficiently handle very large mass spectrometry data and
  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
used for data processing and visualisation. \textcolor{red}{An example
  of its ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in
  1,182 mzXML files) is provided as supplementary information.}  We
will also illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable
\textcolor{red}{and elegant} pipelines.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
  data from an mzML file and} storing \sout{data}\textcolor{red}{them}
either in-memory or on-disk.

@@

\begin{figure}[p]
  \centering
<<plot_bench>>=
(p_read + p_sz + p_filt) / p_access +
    plot_layout(heights = c(1, 0.75)) +
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10))
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
    the benchmarking is a TMT 4-plex experiment acquired on a LTQ
    Orbitrap Velos ((Thermo Fisher Scientific) available in the
    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
\label{fig:bench}
\end{figure}

@@

\textcolor{red}{Because the on-disk backend does not hold all the data
  in memory, direct data manipulations are not possible. We thus
  implemented a \textit{lazy processing} mechanism for this backend
  that caches any data manipulation operations in a processing queue
  in the object itself. Only when the user accesses m/z or intensity
  values are these operations are applied}.  \sout{The on-disk backend
  also offers efficient data manipulation by way of \textit{lazy
    processing} in the object itself.} Operations on the raw data are
stored in a processing queue and only effectively applied when the
user accesses m/z or intensity values. \textcolor{red}{As an
  additional advantage, operations on subsets of the data become much
  faster since data manipulations are applied only to data subsets
  instead of the full data set at once. Also, data access in on-disk
  backends is parallelized by data file ensuring a higher performance
  of this backend over conventional in-memory data representations.}
As an example, the following short analysis pipeline, that can equally
be applied to in-memory or on-disk data, retains MS2 spectra acquired
between 1000 and 3000 seconds, extracts the m/z range corresponding to
the TMT 6-plex range and focuses on the MS2 spectra with a precursor
intensity greater than $11 \times 10^6$ (the median precursor
intensity).

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by retrieving only spectra data from the
selected subset hence avoiding access to the full raw data. Once
access to the spectra m/z and intensity values become mandatory (for
example for plotting), then the in-memory backend becomes more
efficient, as illustrated on Figure~\ref{fig:bench} (d).
\textcolor{red}{The benefit of accessing data in memory is however
  reduced by underlying copies that are performed during the
  subsetting operation. When subsetting an in-memory \texttt{MSnExp}
  into a new, smaller in-memory \texttt{MSnExp} instance, the matrices
  that contain the spectra for the new object are copied, thus leading
  to increased execution time and (transient, if the original data is
  replaced) memory usage. Figure~\ref{fig:bench} (d) shows that the
  larger the subset, that smaller the benefits of an in-memory
  backend. The example with the 6103 spectra, corresponding to the
  full data (i.e. all spectra are already in memory and there is no
  memory management overhead) is representative of memory access only
  and constitutes the best case scenario. } The gain of an in-memory
backend is maximal when the whole dataset is accessed \sout{(i.e. all
  spectra are already in memory)} and becomes negligible when large
subsets of the data are requested.

@@

The \texttt{MSnExp} data structure and its interface constitute an
efficient prototyping environment for computational method
development. We illustrate this by demonstrating how to implement the
BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
acquisition aims at improving the detection of intact precursor ions
by distributing the charge capacity over multiple narrow m/z segments
and thus limiting the proportion of highly abundant precursors in each
segment. A full scan is reconstructed by combining the respective
adjacent segments of the BoxCar acquisitions. The
\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
that demonstrates this. The simple method is composed of three steps
is described below and illustrated with code from
\texttt{MSnbaseBoxCar} in the following code chunk.

>>>>

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. We focus on the new \textit{on-disk} infrastructure, that
  allows the handling of large raw mass spectrometry
  experiment\textcolor{red}{s} on commodity hardware and illustrate
  how the package is used for elegant data processing, method
  development, and visualisation.
\end{abstract}

@@

Mass spectrometry is a powerful technology to assay chemical and
biological samples. It is used in routine applications with well
characterised \textcolor{red}{protocols} \textcolor{red}{such as in
  clinical settings}, as well as a development platform,
\textcolor{red}{with the aim} to improve on existing \sout{methods}
\textcolor{red}{protocols} and devise new ones. The complexity and
diversity of mass spectrometry yield \sout{data that is itself}
complex \textcolor{red}{data} \sout{and often times} of considerable
size, that require non trivial processing before producing
interpretable results. \sout{This is particularly relevant, and can}
\textcolor{red}{The complexity and size of these data} constitute a
significant challenge for \sout{method} \textcolor{red}{protocol}
development: in addition to the development of sample processing and
mass spectrometry methods \textcolor{red}{that yield the raw data},
\textcolor{red}{it is essential} \sout{there is a need} to process,
analyse, interpret and assess these new data to demonstrate the
improvement in the technical, analytical and computational workflows.

Practitioners have a diverse catalogue of software tools \sout{to
  explore, process and interpret mass spectrometry data} at their
disposal. These range from low level software libraries that are aimed
at programmers to enable \textcolor{red}{the} development of new
applications, to more user-oriented applications with graphical user
interfaces which provide a more limited set of functionalities to
address a defined scope. Examples of software libraries include
Java-based jmzML~\cite{Cote:2010} or C/C++-based
ProteoWizard~\cite{Chambers:2012}.  \sout{ProteomeDiscoverer (Thermo
  Scientific)} \textcolor{red}{Thermo Scientific Proteome Discoverer
  (Thermo Fisher Scientific)}, MaxQuant~\cite{Cox:2008} and
PeptideShaker~\cite{Vaudel:2015} are among the most widely used
user-centric applications.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata (Figure \ref{fig:raw}). Each spectrum is
composed of m/z values and associated intensities. The metadata are
represented by a \textcolor{red}{single} table with variables along
the columns and each row associated to a spectrum. Among the metadata
available for each spectrum, there are MS level, acquisition number,
retention time, precursor m/z and intensity (for MS level 2 and
above), and many more. \texttt{MSnbase} \textcolor{red}{relies on the
  \texttt{mzR} package~\cite{Chambers:2012} to import raw mass
  spectrometry data from one of the many community-maintained open
  standards formats (mzML, mzXML, mzData or ANDI-MS/netCDF) and}
provides a rich \textcolor{red}{and principled} interface to
manipulate such objects. The code chunk below illustrates such an
object as displayed in the R console and an enumeration of the
metadata fields.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{./figure/raw.png}
  \caption{\textcolor{red}{Schematic representation of what is
      referred to by \textit{raw data}: a collection of mass spectra
      and a table containing spectrum-level annotations along the
      lines. Raw data are imported from one of the many
      community-maintained open standards formats (mzML, mzXML, mzData
      or ANDI-MS/netCDF).} }
  \label{fig:raw}
\end{figure}

@@

In the following sections, we describe \sout{\texttt{MSnbase}'s
  ability to efficiently handle very large mass spectrometry data and
  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
used for data processing and visualisation. \textcolor{red}{An example
  of its ability to also efficiently handle very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in
  1,182 mzXML files) is provided as supplementary information.}  We
will also illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable
\textcolor{red}{and elegant} pipelines.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to \textcolor{red}{import data from an mzML file} to
create two \texttt{MSnExp} objects that store the data either
in memory or on disk.

@@

\begin{figure}[p]
  \centering
<<plot_bench>>=
(p_read + p_sz + p_filt) / p_access +
    plot_layout(heights = c(1, 0.75)) +
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10))
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
    the benchmarking are a TMT 4-plex experiment acquired on a LTQ
    Orbitrap Velos ((Thermo Fisher Scientific) available in the
    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
\label{fig:bench}
\end{figure}

@@

\textcolor{red}{Because the on-disk backend does not hold all the
  spectra data in memory, direct manipulations of these data are not
  possible. We thus implemented a \textit{lazy processing} mechanism
  for this backend that caches any data manipulation operations in a
  processing queue in the object itself. These operations are then
  applied only when the user accesses m/z or intensity values. }
\sout{The on-disk backend also offers efficient data manipulation by
  way of \textit{lazy processing} in the object itself. Operations on
  the raw data are stored in a processing queue and only effectively
  applied when the user accesses m/z or intensity values.}
\textcolor{red}{As an additional advantage, operations on subsets of
  the data become much faster since data manipulations are applied
  only to data subsets instead of the full data set at once. Also,
  on-disk data access is parallelized by data file ensuring a higher
  performance of this backend over conventional in-memory data
  representations.}  As an example, the following short analysis
pipeline, that can equally be applied to in-memory or on-disk data,
retains MS2 spectra acquired between 1000 and 3000 seconds, extracts
the m/z range corresponding to the TMT 6-plex range and focuses on the
MS2 spectra with a precursor intensity greater than $11 \times 10^6$
(the median precursor intensity).

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by retrieving only spectra data from the
selected subset hence avoiding access to the full raw data. Once
access to the spectra m/z and intensity values become mandatory (for
example for plotting), then the in-memory backend becomes more
efficient, as illustrated on Figure~\ref{fig:bench} (d).
\textcolor{red}{The benefit of accessing data in memory is however
  reduced by underlying copies that are performed during the
  subsetting operation. When subsetting an in-memory \texttt{MSnExp}
  into a new, smaller in-memory \texttt{MSnExp} instance, the matrices
  that contain the spectra for the new object are copied, thus leading
  to increased execution time and (transient, if the original data are
  replaced) memory usage. Figure~\ref{fig:bench} (d) shows that the
  larger the subset, the smaller the benefits of an in-memory backend
  become. The example with the 6103 spectra, corresponding to the full
  data (i.e. all spectra are already in memory and there is no memory
  management overhead) is representative of memory access only and
  constitutes the best case scenario. } \sout{The gain of an in-memory
  backend is maximal when the whole dataset is accessed (i.e. all
  spectra are already in memory) and becomes negligible when large
  subsets of the data are requested.}

@@

The \texttt{MSnExp} data structure and its interface constitute an
efficient prototyping environment for computational method
development. We illustrate this by demonstrating how to implement the
BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
acquisition aims at improving the detection of intact precursor ions
by distributing the charge capacity over multiple narrow m/z segments
and thus limiting the proportion of highly abundant precursors in each
segment. A full scan is reconstructed by combining the respective
adjacent segments of the BoxCar acquisitions. The
\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
that demonstrates this. The simple \sout{method}
\textcolor{red}{pipeline} is composed of three steps,\sout{is}
described below, and illustrated with code from \texttt{MSnbaseBoxCar}
in the following code chunk.

### 2020-07-24 msnbase2.tex
further updates/cleaning of text

\end{knitrout}
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
    the benchmarking is a TMT 4-plex experiment acquired on a LTQ
    Orbitrap Velos ((Thermo Fisher Scientific) available in the
    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
\label{fig:bench}
\end{figure}

>>>>

\end{knitrout}
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
    the benchmarking are a TMT 4-plex experiment acquired on a LTQ
    Orbitrap Velos ((Thermo Fisher Scientific) available in the
    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
\label{fig:bench}
\end{figure}
